Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-18 paper.tex
Updates from review

We have implemented selective sweeps in \msprime\ as a structured
coalescent process~\citep{braverman1995hitchhiking},
following the approach used in \discoal. The first
step in simulating a sweep is to generate
an allele frequency trajectory
from a jump process approximation to the conditional diffusion
of an allele bound for
fixation~\citep{coop2004ancestral}.
The jump process moves back in time following the beneficial
allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
the origination of the allele at $p=1/(2N)$, tracking time in small
increments $\delta t$. Then, given the frequency $p$ at time $t$,
the frequency $p^\prime$ at time $t + \delta t$ is given by
\begin{equation*}
p^\prime =
\begin{cases}
    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
\end{cases}
\end{equation*}
where
\begin{eqnarray*}
	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
\end{eqnarray*}
Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
can be generalised straightforwardly to include arbitrary dominance.
We can also define trajectories to model nuetral alleles and soft
selective sweeps, which we plan as future additions to \msprime.

@@

Fig.~\ref{fig-selection-perf} compares the performance of \msprime\
and \discoal\ under a simple sweep model, and shows
that \msprime\ has far better
CPU time and memory performance.
Since our implementation includes the basic infrastructure for the structured
coalescent, adding support for similar situations,
such as inversions~\citep{peischl2013sequential}, should be straightforward.

>>>>

Selective sweeps are implemented in the coalescent as a two
step-process: first generating an allele frequency trajectory,
and then simulating a structured coalescent process conditioned
on that trajectory.
Following \discoal, we generate sweep trajectories in \msprime\ using a
jump process approximation to the conditional
diffusion of an allele bound for
fixation~\citep{coop2004ancestral}.
The jump process moves back in time following the beneficial
allele frequency, $p$, from some initial frequency (e.g., $p=1$) back to
the origination of the allele at $p=1/(2N)$, tracking time in small
increments $\delta t$. Then, given the frequency $p$ at time $t$,
the frequency $p^\prime$ at time $t + \delta t$ is given by
\begin{equation*}
p^\prime =
\begin{cases}
    p + \mu(p)\delta t +\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
    p + \mu(p)\delta t -\sqrt{p(1-p)\delta t} & \mbox{with probability $1/2$}\\
\end{cases}
\end{equation*}
where
\begin{eqnarray*}
	\mu(p)=\frac{\alpha p (1 - p)}{\tanh(\alpha(1-p))}.
\end{eqnarray*}
Here, $\alpha=2Ns$ and $s$ is the fitness advantage in homozygotes. This model
assumes genic selection (i.e., that the dominance coefficient $h=0.5$), but
can be generalised straightforwardly to include arbitrary dominance.
We can also define trajectories to model nuetral alleles and soft
selective sweeps, which we plan as future additions to \msprime.

@@

Fig.~\ref{fig-selection-perf} compares the performance of \msprime\
and \discoal\ under a simple sweep model, and shows
that \msprime\ has far better
CPU time and memory performance.
Since our implementation uses the abstract label system mentioned above,
adding support for similar situations,
such as inversions~\citep{peischl2013sequential}, should be straightforward.
